An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-04-19 rule-ltl.tex
introduced logic macros

%\klaus{We use ELTL (etc) to denote Extended LTL. 
%But QLTL is also an extension of LTL. These logics 
%are all extensions of LTL. Perhaps RLTL (for Rule-based 
%LTL) would have been better ... or something. Not urgent, 
%we can always change that if we want to.}

@@

An important subset of LTL, called here PLTL, allows only past temporal
operators: $\since$, $\ominus$ and the operators derived
from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is sometimes interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
Alternatively,
it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; in this case, each of the prefixes has to satisfy $\varphi$. This later
form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences. 
(When PLTL is interpreted over finite sequences, its
expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.)

@@

Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
to some important properties. As a canonical example, he demonstrated that it is impossible to
express that all the states with even indexes in a sequence
satisfy $p$ (this is
different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states).
Wolper's example property cannot be expressed in PLTL either.
In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
propositions, as shown below.

@@

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary variables appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states. It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second-order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.

Wolper's property can be expressed in QLTL as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}

Restricting QLTL to the past modalities, one obtains the logic QPLTL. QPLTL has the same expressive
power as regular expressions and finite automata.
Equation~(\ref{form0}) can
be rewritten in QPLTL as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.

%Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
%QPLTL property into a deterministic
%automaton.

@@

We introduce now another extension of LTL, which we call ELTL.
%It has a form that puts it, conceptually, in between temporal
%logic and automata: on the one hand, it is possible to simply write
%a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. 
As will be showed later, this extension is
is very natural for runtime verification.

We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$.
%A model for an ELTL formula has truth assignments only over the propositions in $B$.
An ELTL property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_j$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a PLTL property where propositions from $A$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j$ as a {\em rule}. In text, rules
 will be separated from each other by commas. 
\begin{tabbing}
\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ such that \\
 \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
\end{tabbing}

\noindent
Wolper's example can be written in ELTL
as follows:
\begin{equation} 
\label{form3}
\Box ( q \rightarrow p ) \, \mathrm{where} \, 
 ( q := \ominus \neg q)
\end{equation}
where $A = \{ q\}$ and $B = \{ p \}$.
The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
and each {\em even} state will satisfy $p$.

ELTL extends the set of propositional variables with new variables, whose values at a state are
functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in QLTL, where the values assigned to the
auxiliary variables do not have to extend the 
states of the model in a unique way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to help preventing that the values
of the auxiliary variables are conflicting, as in
$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.

\begin{lemma} \label{fourone}
Let $\varphi$ be an ELTL formula
over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
a set of rules $a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$. 
Let $\sigma$ be a model with states over $B$. 
Then there is at most one
model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $

@@

\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
%A PLTL property depends in a state only on the prefix of the model that precedes and includes that state.

\begin{theorem}
\label{theo1}
The expressive power of 
ELTL is the same as QLTL.
\end{theorem}
\noindent {\bf Proof.}
Each ELTL formula $\eta$, as defined above,
is expressible using the following equivalent QLTL formula:

@@

\noindent
For the other direction, one can translate the QLTL property first into
a second-order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton. The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
The details of the
constructions are given in the appendix. \qed

%EPLTL, where the Muller automaton can be
%replaced with a B\"{u}chi automaton,
%where all of its states, except one, sink
%state, are accepting. \qed

\iffalse
The use of the auxiliary variables $a_1 \ldots a_n$ and the
rules of an ELTL specification 
can be viewed as representing a
state machine, in particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an ELTL specification $\eta$ as consisting of two synchronous parts: an LTL property 
(the statement) 
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary variables, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to QLTL.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary variables. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

%\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}
\fi

We define EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$) in ELTL. This results in
a formalism that is equivalent to finite (B\"{u}chi) automata, where all the states except one are accepting and where the non-accepting state is a sink. 
We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

@@

%For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.

\begin{lemma}
\label{simplecase}
The expressive power of 
EPLTL is the same as QPLTL.
%Any safety QPLTL property can be expressed using the following restricted
%ELTL form:
%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
% {j \in \{1, \ldots , n\}}  \] 
% with $p$ being one of the auxiliary variables %$a_j$.
\end{lemma}
In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
{j \in \{1, \ldots , n\}}  \] 
with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a more intuitive specification.
The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}. An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness

\iffalse
\noindent {\bf Proof.} The proof of the Lemma follows from the
axiomatization of QLTL in~\cite{MP}, e.g.,
$\varphi\ \since\ \psi \leftrightarrow \psi \vee ( \varphi \wedge (\varphi\ \since\ \psi))$. \qed

@@

While this Lemma makes the use of the temporal operators
besides $\ominus$ in EPLTL redundant, it is still
convenient to use a combination of
temporal operators when possible, and
auxiliary variables definitions.
\fi

@@

The algorithm for PLTL first presented in \cite{HR} 
is based on the observation that the semantics of the 
past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
in the previous step $i - 1$ of a subformula.
To demonstrate this, we repeat here the definition of the previous-time
operator $\ominus$, and rewrite the definition
of $\since$ in an equivalent form that is
more directly applicable for runtime verification.
\begin{itemize}
\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
$( \sigma , i)  \models \varphi$ and 
$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}
Runtime verification for a PLTL formula $\eta$ exploits the fact that
the semantic definition is recursive in both the structure of the property. Thus, subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, uses two vectors (arrays) of values indexed by subformulas:  $\old$, which summarizes the truth values of the
subformulas for the execution prefix that
ends just {\em before} the new state, and $\current$, for the execution prefix that
ends with the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
\begin{enumerate}
\item Initially, for each subformula $\varphi$
of $\eta$,
%$B ( \varphi , \current ) = \bfalse $.
$\current ( \varphi ) := \false$.

@@

\subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
formula into a deterministic finite automaton, where one distinguished state, $s$, is a sink state.
Then the sequence of states to be checked forms the input for this automaton, and violation is
announced if $s$ is reached.
QPLTL can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

\subsubsection*{Runtime verification for EPLTL.} For EPLTL,  we need to  add to the above
algorithm calculations 
of $\current ( a_j )$ and
$\current ( \varphi_j )$, for each of the rules of
the form $a_j := \varphi_j$ (the corresponding
$\old$ entries will be updated as in line~3 in the above algorithm).

@@

%For PLTL, we can calculate  $\current$, for subformulas,
%bottom up. But
Because the
auxiliary variables can appear recursively in EPLTL rules, the order of calculation is
subtle. To see this, consider, for example, the formula~(\ref{form3}). It contains
the definition $q := \ominus \neg q$. Now, we cannot calculate this
bottom up, as we did for PLTL, since $\current ( q )$ is not computed yet, and we need 
to calculate $\current ( \ominus \neg q )$ in order 
to compute $\current ( q )$.
However, notice that the calculation is not dependent on the
value of $q$ to calculate $\ominus \neg q$; 
in Step~4 above, we have that
$\current ( \ominus \; \varphi ) := \old ( \varphi )$
so  $\current ( \ominus \neg q ) := \old ( \neg q )$.

@@

We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is relations with
last parameter
that is a natural number represented a time instance. So a tuple of
a relation $R$ can be $(``a", 5, ``cbb", 3)$, where
the  value $3$ is the time parameter. The last parameter $i$
does not correspond to modeling physical time. This
is rather used to allow the relations to have
different tuples in different instances of $i$,
corresponding to states, in the propositional temporal
logics.

@@

The formulas of the core FLTL logic are 
defined by the following grammar,where $p$ denotes a relation,
$a$ denotes a constant and $x$ denotes a variable. 
\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
   \bigcirc \varphi \; | \; 
   ( \varphi \; \until \; \varphi ) \; | $ %\\
   $ \ominus \; \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}

@@

\noindent 
Additional operators are defined as in the propositional logic. We define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past operators
($\since$, $\ominus$ and the ones derived from them) 
forms the logic PFLTL.

@@

For an FLTL (PFLTL) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
We will henceforth sometimes abuse notation, and use the same
symbols both for the relations (semantics) and their representation in
the logic (syntax).
The quantification over values of variables here is {\em static} in the sense that they are independent of the state in the execution. We
denote static quantification with $\exists$ and
$\forall$.
%Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
%and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.

@@

We demonstrate that the lack of expressiveness
carries over from LTL (PLTL) to FLTL (PFLTL).

\vspace{1ex}
\noindent {\bf Example 1.}
Let $p$ and $q$ be temporal relations, where their
restrictions to the state $s_i$, $p[i]$ and $q[i]$, are unary relations $p$ and $q$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). To show that this is not expressible in FLTL (and PFLTL),
consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
that in this case, for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.
Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015}, can express this property, 
where for each value $a$ there is a separate automaton that counts the number of times that $q(a)$ has occurred.

>>>>

%\klaus{We use \ELTL{} (etc) to denote Extended LTL. 
%But \QLTL{} is also an extension of LTL. These logics 
%are all extensions of LTL. Perhaps RLTL (for Rule-based 
%LTL) would have been better ... or something. Not urgent, 
%we can always change that if we want to.}

@@

An important subset of LTL, called here \PLTL{}, allows only past temporal
operators: $\since$, $\ominus$ and the operators derived
from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is sometimes interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
Alternatively,
it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; in this case, each of the prefixes has to satisfy $\varphi$. This later
form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences. 
(When \PLTL{} is interpreted over finite sequences, its
expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.)

@@

Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
to some important properties. As a canonical example, he demonstrated that it is impossible to
express that all the states with even indexes in a sequence
satisfy $p$ (this is
different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states).
Wolper's example property cannot be expressed in \PLTL{} either.
In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
propositions, as shown below.

@@

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary variables appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states. It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic \QLTL{} has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second-order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of \QLTL{}~\cite{Thomas}.

Wolper's property can be expressed in \QLTL{} as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}

Restricting \QLTL{} to the past modalities, one obtains the logic \QPLTL{}. \QPLTL{} has the same expressive
power as regular expressions and finite automata.
Equation~(\ref{form0}) can
be rewritten in \QPLTL{} as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.

%Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
%\QPLTL{} property into a deterministic
%automaton.

@@

We introduce now another extension of LTL, which we call \ELTL{}.
%It has a form that puts it, conceptually, in between temporal
%logic and automata: on the one hand, it is possible to simply write
%a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. 
As will be showed later, this extension is
is very natural for runtime verification.

We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$.
%A model for an \ELTL{} formula has truth assignments only over the propositions in $B$.
An \ELTL{} property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_j$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a \PLTL{} property where propositions from $A$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j$ as a {\em rule}. In text, rules
 will be separated from each other by commas. 
\begin{tabbing}
\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ such that \\
 \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
\end{tabbing}

\noindent
Wolper's example can be written in \ELTL{}
as follows:
\begin{equation} 
\label{form3}
\Box ( q \rightarrow p ) \, \mathrm{where} \, 
 ( q := \ominus \neg q)
\end{equation}
where $A = \{ q\}$ and $B = \{ p \}$.
The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
and each {\em even} state will satisfy $p$.

\ELTL{} extends the set of propositional variables with new variables, whose values at a state are
functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in \QLTL{}, where the values assigned to the
auxiliary variables do not have to extend the 
states of the model in a unique way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to help preventing that the values
of the auxiliary variables are conflicting, as in
$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.

\begin{lemma} \label{fourone}
Let $\varphi$ be an \ELTL{} formula
over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
a set of rules $a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$. 
Let $\sigma$ be a model with states over $B$. 
Then there is at most one
model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $

@@

\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
is defined, via a \PLTL{} formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
%A \PLTL{} property depends in a state only on the prefix of the model that precedes and includes that state.

\begin{theorem}
\label{theo1}
The expressive power of 
\ELTL{} is the same as \QLTL{}.
\end{theorem}
\noindent {\bf Proof.}
Each \ELTL{} formula $\eta$, as defined above,
is expressible using the following equivalent \QLTL{} formula:

@@

\noindent
For the other direction, one can translate the \QLTL{} property first into
a second-order monadic logic formula, then to a deterministic Muller automata and then construct an \ELTL{} formula that holds for the accepting executions of this automaton. The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
The details of the
constructions are given in the appendix. \qed

%\EPLTL{}, where the Muller automaton can be
%replaced with a B\"{u}chi automaton,
%where all of its states, except one, sink
%state, are accepting. \qed

\iffalse
The use of the auxiliary variables $a_1 \ldots a_n$ and the
rules of an \ELTL{} specification 
can be viewed as representing a
state machine, in particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an \ELTL{} specification $\eta$ as consisting of two synchronous parts: an LTL property 
(the statement) 
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary variables, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to \QLTL{}.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary variables. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

%\klaus{First sentence below sounds wrong: (1) \PLTL{} already does not have future time operators, and (2) why mention infinite sequences?}
\fi

We define \EPLTL{} by disallowing the future temporal operators (except an implied prefixing $\Box$) in \ELTL{}. This results in
a formalism that is equivalent to finite (B\"{u}chi) automata, where all the states except one are accepting and where the non-accepting state is a sink. 
We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

@@

%For representing safety properties that can be expressed in \QPLTL{}, we can use a restricted version as shown in the following simple Lemma.

\begin{lemma}
\label{simplecase}
The expressive power of 
\EPLTL{} is the same as \QPLTL{}.
%Any safety \QPLTL{} property can be expressed using the following restricted
%\ELTL{} form:
%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
% {j \in \{1, \ldots , n\}}  \] 
% with $p$ being one of the auxiliary variables %$a_j$.
\end{lemma}
In fact, it is easy to show that \EPLTL{} can be restricted to the following form, while retaining its expressive power.
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
{j \in \{1, \ldots , n\}}  \] 
with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
of an \EPLTL{} property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a more intuitive specification.
The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}. An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness

\iffalse
\noindent {\bf Proof.} The proof of the Lemma follows from the
axiomatization of \QLTL{} in~\cite{MP}, e.g.,
$\varphi\ \since\ \psi \leftrightarrow \psi \vee ( \varphi \wedge (\varphi\ \since\ \psi))$. \qed

@@

While this Lemma makes the use of the temporal operators
besides $\ominus$ in \EPLTL{} redundant, it is still
convenient to use a combination of
temporal operators when possible, and
auxiliary variables definitions.
\fi

@@

The algorithm for \PLTL{} first presented in \cite{HR} 
is based on the observation that the semantics of the 
past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
in the previous step $i - 1$ of a subformula.
To demonstrate this, we repeat here the definition of the previous-time
operator $\ominus$, and rewrite the definition
of $\since$ in an equivalent form that is
more directly applicable for runtime verification.
\begin{itemize}
\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
$( \sigma , i)  \models \varphi$ and 
$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}
Runtime verification for a \PLTL{} formula $\eta$ exploits the fact that
the semantic definition is recursive in both the structure of the property. Thus, subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, uses two vectors (arrays) of values indexed by subformulas:  $\old$, which summarizes the truth values of the
subformulas for the execution prefix that
ends just {\em before} the new state, and $\current$, for the execution prefix that
ends with the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
\begin{enumerate}
\item Initially, for each subformula $\varphi$
of $\eta$,
%$B ( \varphi , \current ) = \bfalse $.
$\current ( \varphi ) := \false$.

@@

\subsubsection*{Runtime verification for \QPLTL{}.}  This can be performed by translating a
formula into a deterministic finite automaton, where one distinguished state, $s$, is a sink state.
Then the sequence of states to be checked forms the input for this automaton, and violation is
announced if $s$ is reached.
\QPLTL{} can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

\subsubsection*{Runtime verification for \EPLTL{}.} For \EPLTL{},  we need to  add to the above
algorithm calculations 
of $\current ( a_j )$ and
$\current ( \varphi_j )$, for each of the rules of
the form $a_j := \varphi_j$ (the corresponding
$\old$ entries will be updated as in line~3 in the above algorithm).

@@

%For \PLTL{}, we can calculate  $\current$, for subformulas,
%bottom up. But
Because the
auxiliary variables can appear recursively in \EPLTL{} rules, the order of calculation is
subtle. To see this, consider, for example, the formula~(\ref{form3}). It contains
the definition $q := \ominus \neg q$. Now, we cannot calculate this
bottom up, as we did for \PLTL{}, since $\current ( q )$ is not computed yet, and we need 
to calculate $\current ( \ominus \neg q )$ in order 
to compute $\current ( q )$.
However, notice that the calculation is not dependent on the
value of $q$ to calculate $\ominus \neg q$; 
in Step~4 above, we have that
$\current ( \ominus \; \varphi ) := \old ( \varphi )$
so  $\current ( \ominus \neg q ) := \old ( \neg q )$.

@@

We define models for \FLTL{} based on {\em temporal} relations~\cite{Chomicki}. That is relations with
last parameter
that is a natural number represented a time instance. So a tuple of
a relation $R$ can be $(``a", 5, ``cbb", 3)$, where
the  value $3$ is the time parameter. The last parameter $i$
does not correspond to modeling physical time. This
is rather used to allow the relations to have
different tuples in different instances of $i$,
corresponding to states, in the propositional temporal
logics.

@@

The formulas of the core \FLTL{} logic are 
defined by the following grammar,where $p$ denotes a relation,
$a$ denotes a constant and $x$ denotes a variable. 
\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
   \bigcirc \varphi \; | \; 
   ( \varphi \; \until \; \varphi ) \; | $ %\\
   $ \ominus \; \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}

@@

\noindent 
Additional operators are defined as in the propositional logic. We define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past operators
($\since$, $\ominus$ and the ones derived from them) 
forms the logic \PFLTL{}.

@@

For an \FLTL{} (\PFLTL{}) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
We will henceforth sometimes abuse notation, and use the same
symbols both for the relations (semantics) and their representation in
the logic (syntax).
The quantification over values of variables here is {\em static} in the sense that they are independent of the state in the execution. We
denote static quantification with $\exists$ and
$\forall$.
%Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
%and will extend \FLTL{} below, is expressed with $\Exists$ and $\Forall$.

@@

We demonstrate that the lack of expressiveness
carries over from LTL (\PLTL{}) to \FLTL{} (\PFLTL{}).

\vspace{1ex}
\noindent {\bf Example 1.}
Let $p$ and $q$ be temporal relations, where their
restrictions to the state $s_i$, $p[i]$ and $q[i]$, are unary relations $p$ and $q$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). To show that this is not expressible in \FLTL{} (and \PFLTL{}),
consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
that in this case, for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.
Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015}, can express this property, 
where for each value $a$ there is a separate automaton that counts the number of times that $q(a)$ has occurred.

\vspace{1ex}
\noindent {\bf Example 2.}
Consider the property that asserts that when
$\mathit{report}(y , x, d)$ appears
in a state, denoting that  process 
$y$ sends some data $d$ to a process $x$,
there was a chain of states with process spawns
$\mathit{spawn}(x, x_1)$, $\mathit{spawn} (x_1, x_2)$ $\ldots$
$\mathit{spawn}(x_l , y)$. i.,e., $y$ is a descendent process of $x$. The reason
is that the required property needs to
assert about the transitive closure
of the relation $\mathit{spawn}$. 
\FLTL{} can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}. Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, ((p(x, t) \rightarrow \exists t' p(x, t') )$\footnote{The obtained model will have, in addition
to the aforementioned relations, also
the linear order $<$ over the naturals.}.
It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in first order setting. The proof is based on the
compactness theory of first order logic~\cite{Flum}.

\subsubsection{Extending \FLTL{} with dynamic quantification.}
